Type-Safe Modular Parsing

### 2017-01-13 paper/paper.tex

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based parsing combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 68\% shorter than the non-modular
  parsing code.

>>>>

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based parsing combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{multiple inheritance} and 
  \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 69\% shorter than an existing code
  based using non-modular parsing code.

### 2017-01-13 paper/sections/Conclusion.tex

Huang: to write a couple of sentences for a summary.

\noindent Haoyuan: to discuss future work. (some boilerplate can be avoided via metaprogramming.)

>>>>

\haoyuan{Huang: to write a couple of sentences for a summary.}

For a better user experience, there are certainly some aspects in our pattern for modular parsing that needs
improvements. Specifically, in Section~\ref{subsec:implementation} we have observed that the glue code in \lstinline{Untyped}
appears to be boilerplate, as we need to compose the data structures, parsers and operations respectively from the same set of super types. Such an issue refers to \textit{family polymorphism} [], and we may rely it on a language feature or meta-programming techniques. Moreover, to make better use of Object Algebras, we can possibly adopt some patterns in Shy, together with the composition of algebras, to reduce boilerplate by using code generation.

For future work, we will experiment more on how open recursion contributes to extensible parsing in functional languages, by making use of laziness. It is also interesting to see that parsing, to some extent, can be viewed as a special example of unfolds. So it is worthwhile considering to generalize the composition of unfolds under certain circumstances.

### 2017-01-13 paper/sections/Discussion.tex

\begin{figure}[t]
\begin{tabular}{m{0.42\linewidth}m{0.52\linewidth}}
\lstinputlisting[linerange=113-128]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
&
\lstinputlisting[linerange=132-147]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\end{tabular}
\caption{Language \lstinline{VarExpr} (left) and language \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\lstinputlisting[linerange=151-160]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}[]. Unfortunately Scala does not
support this feature, nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

>>>>

\begin{figure}[t]
\begin{tabular}{m{0.42\linewidth}m{0.52\linewidth}}
\lstinputlisting[linerange=113-132]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
&
\lstinputlisting[linerange=136-156]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\end{tabular}
\caption{Language \lstinline{VarExpr} (left) and language \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}[]. Unfortunately Scala does not
support this feature, nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

### 2017-01-13 paper/sections/Introduction.tex

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice. Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. As Kastner~\cite{} notes, the typical
drawback of such techniques is that
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.\bruno{mention that a reason why syntactic 
modularization techniques are popular is simplicity (in implementation
and also use).}

Semantic modularization techniques move away
from syntactic composition techniques that rely on the textual source
code. This allows us to go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a Software
Engineering point-of-view, and enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{}, as 
well as several techniques for solving the Expression Problem. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.} Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

>>>>

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use. 
Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. As Kastner~\cite{} notes, the typical
drawback of such techniques is that
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Furthermore they enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{},
virtual classes~\cite{}, as 
well as several other techniques for solving the Expression Problem~\cite{}. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.}

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the perceived need
for sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras are one such technique, which 
Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.
